### 2025-03-31 README.Rmd
update documentation.

The goal of baseISS is to develop a preliminary method for calculating input sample size (ISS) for fishery-dependent composition data. This data includes length- and age-frequency data collected by the NORPAC observer program and excludes fishery-independent data from surveys. Fishery-dependent composition data represents an important element of the 26 statistical catch at age stock assessment models for Alaskan fisheries managed by the NOAA AFSC. The ISS determines the precision attached to this data type, and therefore regulates its leverage in assessment model outputs such as biomass predictions.

The structure and functionality of this package resembles surveyISS with some notable differences that accomodate the neuances of fishery-dependent data. These are covered in the next section, "Important Notes".

## Important Notes

### Input Data
The following conditions of an input data frame for any species must be met:
1. (length data) There exists a WEIGHT1 column that represents the haul-level proportions-at-length of the observed lengths. This is calculated by dividing the frequency of the observed length within a haul by the total number of fish lengthed in that haul. It should be a standard calculation that is exactly the same for every species. This convention...

2. Length data be condensed with sum_frequency column.

3. Age data include associated lengths.

4. Age data flattened, no "sum_frequency" column.

@@

1. Length ISS must be a separate analysis to age ISS. Unlike surveyISS, this is necessary for several reasons: a. there will be many more hauls with lengths than ages, b. it is likely that the user options such as post_stratify will be different between length data and age data, c. NOTE ANOTHER ONE WHEN YOU REMEMBER.

### Developing Functionality

## Installation

@@

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(baseISS)
## basic example code
```

What is special about using `README.Rmd` instead of just `README.md`? You can include R chunks like so:

```{r cars}
summary(cars)
```

You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this.

You can also embed plots, for example:

```{r pressure, echo = FALSE}
plot(pressure)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.

>>>>

The goal of baseISS is to develop a preliminary bootstrap method for calculating input sample size (ISS) for fishery-dependent composition data. This data includes length- and age-frequency data collected by the NORPAC observer program and excludes fishery-independent data from surveys. Fishery-dependent composition data represents an important element of the 26 statistical catch at age stock assessment models for Alaskan fisheries managed by the NOAA AFSC. The ISS determines the precision attached to this data type, and therefore regulates (to a degree) its leverage in assessment model outputs such as biomass predictions.

The structure and functionality of this package resembles surveyISS with some notable differences that accommodate the nuances of fishery-dependent data.As of writing this, baseISS can perform the bootstrap ISS calculation for EBS Pcod length data, and has been successfully tested on EBS Pcod age data as well as Pollock length data.

### Process Flow

fishery_iss.R is the primary function that controls all other processes to calculate ISS. The descriptions of the function arguments in fishery_iss.R are the key resource for its capability. There are many options a user can toggle to suit their particular requirements, such as bootstrapping trips or hauls instead of length or age observations.

This is an example of specifying fishery_iss.R for EBS Pcod:
```{r, eval=F}

fishery_iss(species_code = "202",
            area_code = "EBS",
            length_based = TRUE,
            iters = iterations,
            freq_data = lfreq_data,
            yrs = NULL,
            bin = NULL,
            plus_len = NULL,
            post_strata = NULL,
            minimum_sample_size = list(resolution = "YAGM_SFREQ", size = 30),
            new_trip_N = NULL,
            new_haul_N = NULL,
            new_length_N = NULL,
            max_length = 150,
            boot.trip = TRUE,
            boot.haul = TRUE,
            boot.length = TRUE,
            boot.age = FALSE,
            expand.by.sampling.strata = FALSE,
            expansion_factors = c("haul_numbers", "month_numbers"),
            save_data_frame = TRUE) -> .output

```

### Input Data
The following conditions of an input data frame for any species must be met:

1. Age data include associated lengths.

2. Age data needs to be flattened, meaning every observation has a row in the data frame whether it is identical to another or not. This ensures that the length measurement associated with the age measurement is preserved.

3. For EBS Pcod, age data has only been used when there exists already pre-processed length data associated with it. I.e., I matched the y2 length data object received from Steve with age data I downloaded when testing the baseISS functionality on age data. This was necessary because I needed quantities like YAGM_SNUM to perform the weighted expansion.

@@

1. Length ISS must be a separate analysis to age ISS. Unlike surveyISS, this is necessary for several reasons: a. there will be many more hauls with lengths than ages for fishery-dependent data, b. it is likely that the user options such as post_stratify will be different between length data and age data, and these must be performed separately between length and age analyses.

2. You can't chose the species from a big data set containing all the species like you can in surveyISS. This is because data sets for each species should include unique columns necessary to apply their unique expansion methods. For EBS Pcod for example, these columns are only available after much pre-processing of the raw data. The user can input a species-specific data frame into fishery_iss.R and run it according to their specifications.

3. A user can add their species- and area-specific expansion routine to expand_props.R. As of writing this, the only official expansion routine coded is for EBS Pcod length data from the 2023 assessment. There are also placeholder expansion routines for age data for the EBS Pcod and length data for Pollock to test the generic functionality of baseISS.

### Developing Functionality
- different species - tested with Pollock only. Need to re-test with pollock using generic data download.
- reboot_age.R - needs developing
- combine post-strata. needs developing.

### Installation

@@

### Example
Coming Soon...
